
This analysis examines projected *cross-border migration* dynamics across Africa under future climate and socioeconomic scenarios (SSPs) for 2020 and 2050. Using country-level migration data, we quantify inflows and outflows, identify top destination and origin countries, and aggregate flows by African sub-regions. We complement these analyses with country-level composition maps and bilateral migration chord diagrams to reveal structural connectivity and dominant migration corridors. 

Part 0: Libraries & Setup
```{r}
# Libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
library(readr)
library(sf)
library(scatterpie)
library(ggrepel)
library(RColorBrewer)
library(stringr)
library(circlize)

# Working directory
setwd("/Users/drdesalewmoges/Documents/Papers/Climate_dengue")

# Load main dataset
data <- read_csv(
  "Data/BilateralMigration/Summary_ACMI_bilateral_migration_2015_2050.csv"
)
```

Part 1: Continental Cumulative Migration by Scenario
```{r}
flows_long <- data %>%
  select(
    dest_iso3c,
    orig_iso3c,
    year,
    SSP126 = GEPIC_WG2_SSP1_RCP26D_mean,
    SSP160 = GEPIC_WG2_SSP1_RCP60D_mean,
    SSP360 = GEPIC_WG2_SSP3_RCP60D_mean
  ) %>%
  pivot_longer(
    cols = c(SSP126, SSP160, SSP360),
    names_to = "scenario",
    values_to = "flow"
  ) %>%
  filter(year != 2015)

agg_inflows <- flows_long %>%
  group_by(year, scenario) %>%
  summarise(
    total_inflow = sum(abs(flow), na.rm = TRUE),
    .groups = "drop"
  )

agg_inflows_cum <- agg_inflows %>%
  arrange(scenario, year) %>%
  group_by(scenario) %>%
  mutate(cumulative_inflow = cumsum(total_inflow)) %>%
  ungroup()

agg_inflows_cum$scenario <- factor(
  agg_inflows_cum$scenario,
  levels = c("SSP126", "SSP160", "SSP360"),
  labels = c("SSP1-2.6", "SSP1-6.0", "SSP3-6.0")
)

plot_cumulative <- ggplot(
  agg_inflows_cum,
  aes(x = as.numeric(year), y = cumulative_inflow / 1e3, color = scenario)
) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = comma) +
  scale_color_brewer(palette = "Set2") +
  labs(
    x = "Year",
    y = "Cumulative Migrants (Thousands)",
    color = "Scenario"
  ) +
  theme_bw()

plot_cumulative

ggsave(
  "Output/plot_cumulative.pdf",
  plot_cumulative,
  width = 4,
  height = 3,
  dpi = 300
)
dev.off()
```

Part 2: Top 20 Destination Countries (2020–2050) with Net Change Overlay
```{r}
# Scenario column
ssp_col <- "GEPIC_WG2_SSP1_RCP60D_mean"

# Function: Prepare inflow data
prepare_flow <- function(df, top_n = 20) {

  # Inflows in 2020 and 2050
  df_summary <- df %>%
    filter(year %in% c(2020, 2050)) %>%
    group_by(dest_iso3c, year) %>%
    summarise(
      total = sum(abs(mean), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    pivot_wider(
      names_from  = year,
      values_from = total,
      names_prefix = "y"
    ) %>%
    mutate(
      y2020  = round(y2020, 0),
      y2050  = round(y2050, 0),
      change = y2050 - y2020
    )
  # Total inflow over full period (2020–2050)
  df_total <- df %>%
    filter(year >= 2020, year <= 2050) %>%
    group_by(dest_iso3c) %>%
    summarise(
      total_inflow = round(sum(abs(mean), na.rm = TRUE), 0),
      .groups = "drop"
    )

  df_summary %>%
    left_join(df_total, by = "dest_iso3c") %>%
    arrange(desc(total_inflow)) %>%
    slice_head(n = top_n)
}

# Function: Plot inflows with change overlay
plot_inflow <- function(df, scenario_name = "SSP1-6.0") {

  ggplot(
    df,
    aes(y = reorder(dest_iso3c, total_inflow))
  ) +
    # Base bar: cumulative inflow (2020–2050)
    geom_col(
      aes(x = total_inflow),
      fill = "#CDCD00"
    ) +
    # Overlay rectangle: change from 2020 to 2050
    geom_rect(
      aes(
        xmin = pmin(total_inflow, total_inflow + change),
        xmax = pmax(total_inflow, total_inflow + change),
        ymin = as.numeric(reorder(dest_iso3c, total_inflow)) - 0.3,
        ymax = as.numeric(reorder(dest_iso3c, total_inflow)) + 0.3,
        fill = change > 0
      ),
      alpha = 0.7
    ) +
    # Change value labels
    geom_text(
      aes(
        x = total_inflow + change,
        label = comma(change),
        hjust = if_else(change > 0, -0.15, -0.4)
      ),
      size = 3,
      colour = "black"
    ) +
    # Color scale for change
    scale_fill_manual(
      values = c(
        "TRUE"  = "#EE0000",
        "FALSE" = "#008B45"
      ),
      labels = c(
        "Decrease from 2020 to 2050",
        "Increase from 2020 to 2050"
      ),
      name = "Change in inflow"
    ) +
    # Axis scaling
    scale_x_continuous(
      labels = comma,
      expand = expansion(mult = c(0.05, 0.15))
    ) +
    # Labels
    labs(
      x = "Total inflow (2020–2050)",
      y = "Destination country",
      title = scenario_name
    ) +
    # Theme
    theme_bw(base_size = 12) +
    theme(
      panel.grid      = element_blank(),
      legend.position = c(0.8, 0.3),
      legend.title    = element_text(size = 7, face = "bold"),
      legend.text     = element_text(size = 7),
      axis.title      = element_text(size = 7, face = "bold"),
      axis.text       = element_text(size = 7, face = "bold")
    )
}

# Prepare data for SSP1-6.0
df_scenario <- data %>%
  select(
    orig_iso3c,
    dest_iso3c,
    year,
    mean = all_of(ssp_col)
  )

inflow_df <- prepare_flow(df_scenario, top_n = 20)

# Generate plot
ssp160_inflow_plot <- plot_inflow(
  inflow_df,
  scenario_name = "SSP1-6.0"
)
ssp160_inflow_plot

# Save figure
ggsave(
  filename = "Output/SSP160_Top20_Inflow_2020_2050.pdf",
  plot = ssp160_inflow_plot,
  width = 5,
  height = 3,
  dpi = 300
)
dev.off()
```

Part 3: Country-Level Inflow vs Outflow Proportions (2020 & 2050)
```{r}
# Scenario column
ssp_col <- "GEPIC_WG2_SSP1_RCP60D_mean"

# Read Africa shapefile
africa <- read_sf("Data/Africa_regions.gpkg") %>%
  filter(!ISO3 %in% c("CPV", "STP", "SYC", "MUS", "REU", "COM")) %>%
  select(ISO3, region, NAME, geom)

# Compute country centroids
africa_centroids <- africa %>%
  st_point_on_surface() %>%
  mutate(
    x = st_coordinates(.)[, 1],
    y = st_coordinates(.)[, 2]
  ) %>%
  select(ISO3, NAME, x, y)

# Prepare migration data (2020 & 2050)
df_scen <- data %>%
  select(orig_iso3c, dest_iso3c, year, mean = all_of(ssp_col)) %>%
  filter(year %in% c(2020, 2050))

# ---- Inflows ----
inflow <- df_scen %>%
  group_by(dest_iso3c, year) %>%
  summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
  rename(ISO3 = dest_iso3c) %>%
  mutate(type = "Inflow")

# ---- Outflows ----
outflow <- df_scen %>%
  group_by(orig_iso3c, year) %>%
  summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
  rename(ISO3 = orig_iso3c) %>%
  mutate(type = "Outflow")

# Combine and attach coordinates
bubbles_df <- bind_rows(inflow, outflow) %>%
  left_join(africa_centroids, by = "ISO3") %>%
  filter(!is.na(x) & !is.na(value))

# Pivot to pie format & compute radius
pie_df <- bubbles_df %>%
  select(ISO3, year, type, value, x, y) %>%
  pivot_wider(
    names_from = type,
    values_from = value,
    values_fill = 0
  ) %>%
  mutate(
    total = Inflow + Outflow,
    r = rescale(total, to = c(3, 8))
  )

# Join country metadata (adds region)
pie_country <- pie_df %>%
  left_join(africa %>% st_set_geometry(NULL), by = "ISO3")

# Conditional jittering
regions_to_jitter <- c("West Africa", "Southern Africa")

pie_country <- pie_country %>%
  group_by(year) %>%
  mutate(
    do_jitter = region %in% regions_to_jitter,
    jitter_scale = if_else(do_jitter, rescale(r, to = c(1.5, 2.5)), 0),
    x_plot = x + runif(n(), -3, 3) * jitter_scale,
    y_plot = y + runif(n(), -3, 3) * jitter_scale
  ) %>%
  ungroup()

# Plot
gg_country <- ggplot() +

  # ---- Base map ----
  geom_sf(
    data = africa,
    fill = "#C7C7C7",
    color = "#FFFFFF",
    linewidth = 0.2
  ) +

  # ---- Leader lines (only jittered countries) ----
  geom_segment(
    data = pie_country %>% filter(do_jitter),
    aes(x = x, y = y, xend = x_plot, yend = y_plot),
    colour = "#E0E0E0",
    linewidth = 0.25
  ) +

  # ---- Scatter pies ----
  geom_scatterpie(
    data = pie_country,
    aes(x = x_plot, y = y_plot, r = r),
    cols = c("Inflow", "Outflow"),
    alpha = 0.75
  ) +

  # ---- Country labels (ISO3) ----
  geom_text(
    data = pie_country,
    aes(x = x_plot, y = y_plot, label = ISO3),
    size = 2,
    colour = "black",
    face = "bold"
  ) +

  # ---- Facet by year ----
  facet_wrap(~year) +

  # ---- Colors ----
  scale_fill_manual(
    values = c(
      "Inflow"  = "#76EEC6",
      "Outflow" = "#FF7F50"
    ),
    name = "Flow type\n(proportions)"
  ) +

  coord_sf() +
  theme_void() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(size = 12, face = "bold")
  )

# Show plot
gg_country

# Save figure
ggsave(
  filename = "Output/Country_PieFlows_2020_2050_SSP160.pdf",
  plot = gg_country,
  width = 10,
  height = 6,
  dpi = 300
)
dev.off()
```

Part 4: Regional Migration Totals (2020 vs 2050)
```{r}
# Scenario column
ssp_col <- "GEPIC_WG2_SSP1_RCP60D_mean"

# Read Africa shapefile (for regions)
africa <- read_sf("Data/Africa_regions.gpkg") %>%
  select(ISO3, region) %>%
  st_set_geometry(NULL)

# Prepare migration data (2020 & 2050)
df_scen <- data %>%
  select(orig_iso3c, dest_iso3c, year, mean = all_of(ssp_col)) %>%
  filter(year %in% c(2020, 2050))

# Inflows by destination region
regional_inflow <- df_scen %>%
  group_by(dest_iso3c, year) %>%
  summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
  rename(ISO3 = dest_iso3c) %>%
  left_join(africa, by = "ISO3") %>%
  group_by(region, year) %>%
  summarise(inflow = sum(value, na.rm = TRUE), .groups = "drop")

# Outflows by origin region
regional_outflow <- df_scen %>%
  group_by(orig_iso3c, year) %>%
  summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
  rename(ISO3 = orig_iso3c) %>%
  left_join(africa, by = "ISO3") %>%
  group_by(region, year) %>%
  summarise(outflow = sum(value, na.rm = TRUE), .groups = "drop")

# Combine inflow & outflow
regional_flows <- full_join(
  regional_inflow,
  regional_outflow,
  by = c("region", "year")
) %>%
  replace_na(list(inflow = 0, outflow = 0)) %>%
  pivot_longer(
    cols = c(inflow, outflow),
    names_to = "type",
    values_to = "value"
  )

# Plot
gg_region <- ggplot(
  regional_flows,
  aes(
    x = region,
    y = value,
    fill = type
  )
) +

  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +

  facet_wrap(~year) +

  scale_fill_manual(
    values = c(
      "inflow"  = "#76EEC6",
      "outflow" = "#FF7F50"
    ),
    labels = c(
      "Inflow",
      "Outflow"
    ),
    name = "Migration flow"
  ) +

  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.1))
  ) +

  labs(
    x = "African region",
    y = "Total migration",
    title = "Regional Migration Totals",
    subtitle = "Comparison between 2020 and 2050 (SSP1–6.0)"
  ) +

  theme_bw(base_size = 11) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.x = element_text(angle = 30, hjust = 1, face = "bold"),
    axis.title  = element_text(face = "bold"),
    strip.text  = element_text(size = 12, face = "bold"),
    legend.position = "bottom"
  )

# Show plot
gg_region

# Save figure
ggsave(
  filename = "Output/Regional_Migration_Totals_2020_2050_SSP160.pdf",
  plot = gg_region,
  width = 9,
  height = 5,
  dpi = 300
)
dev.off()
```

Part 5 (Supplementary): Top 10 Inflows & Outflows by Scenario
```{r}
ssp_cols <- c(
  "SSP1-6.0" = "GEPIC_WG2_SSP1_RCP60D_mean",
  "SSP2-4.5" = "GEPIC_WG2_SSP2_RCP45D_mean",
  "SSP3-7.0" = "GEPIC_WG2_SSP3_RCP70D_mean"
)

# Years of interest
years_use <- c(2020, 2050)

# Function: prepare top 10 flows
prepare_top10 <- function(df, flow_type = c("inflow", "outflow")) {

  flow_type <- match.arg(flow_type)

  if (flow_type == "inflow") {
    df %>%
      group_by(dest_iso3c, year, scenario) %>%
      summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
      rename(country = dest_iso3c)
  } else {
    df %>%
      group_by(orig_iso3c, year, scenario) %>%
      summarise(value = sum(abs(mean), na.rm = TRUE), .groups = "drop") %>%
      rename(country = orig_iso3c)
  }
}

# Build dataset for all SSPs
top10_data <- map_dfr(names(ssp_cols), function(ssp) {

  data %>%
    select(orig_iso3c, dest_iso3c, year,
           mean = all_of(ssp_cols[[ssp]])) %>%
    filter(year %in% years_use) %>%
    mutate(scenario = ssp)
})

# Inflows & outflows
inflow_top10 <- prepare_top10(top10_data, "inflow")
outflow_top10 <- prepare_top10(top10_data, "outflow")


# Select top 10 per scenario/year
top10_combined <- bind_rows(
  inflow_top10 %>% mutate(type = "Inflow"),
  outflow_top10 %>% mutate(type = "Outflow")
) %>%
  group_by(scenario, year, type) %>%
  slice_max(order_by = value, n = 10) %>%
  ungroup()

# Plot
gg_top10 <- ggplot(
  top10_combined,
  aes(
    x = reorder(country, value),
    y = value,
    fill = type
  )
) +

  geom_col(width = 0.7) +
  coord_flip() +

  facet_grid(
    scenario ~ year,
    scales = "free_y"
  ) +

  scale_fill_manual(
    values = c(
      "Inflow"  = "#76EEC6",
      "Outflow" = "#FF7F50"
    )
  ) +

  scale_y_continuous(labels = scales::comma) +

  labs(
    x = NULL,
    y = "Migration volume",
    title = "Top 10 Inflows and Outflows by Scenario",
    subtitle = "2020 vs 2050"
  ) +

  theme_bw(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    strip.text = element_text(size = 10, face = "bold"),
    legend.position = "bottom"
  )

# Show plot
gg_top10

# Save figure
ggsave(
  filename = "Output/Supp_Top10_Flows_By_Scenario.pdf",
  plot = gg_top10,
  width = 12,
  height = 8,
  dpi = 300
)
dev.off()
```

Part 6 (Supplementary): Bilateral Migration Chord Diagrams
```{r}
ssp_col <- "GEPIC_WG2_SSP1_RCP60D_mean"
year_use <- 2050

# Prepare bilateral flows
bilateral <- data %>%
  select(orig_iso3c, dest_iso3c, year,
         mean = all_of(ssp_col)) %>%
  filter(year == year_use) %>%
  mutate(flow = abs(mean)) %>%
  group_by(orig_iso3c, dest_iso3c) %>%
  summarise(value = sum(flow, na.rm = TRUE), .groups = "drop") %>%
  filter(value > quantile(value, 0.90, na.rm = TRUE))  # top 10% only

# Convert to matrix
flow_matrix <- bilateral %>%
  pivot_wider(
    names_from = dest_iso3c,
    values_from = value,
    values_fill = 0
  ) %>%
  column_to_rownames("orig_iso3c") %>%
  as.matrix()

# Color palette
country_colors <- setNames(
  colorRampPalette(c("#2166AC", "#B2182B"))(nrow(flow_matrix)),
  rownames(flow_matrix)
)

# Plot
circos.clear()
circos.par(start.degree = 90, gap.degree = 5)

chordDiagram(
  x = flow_matrix,
  grid.col = country_colors,
  transparency = 0.25,
  directional = 1,
  direction.type = c("arrows", "diffHeight"),
  link.arr.type = "big.arrow"
)

title(
  main = "Dominant Bilateral Migration Corridors (2050, SSP1–6.0)",
  cex.main = 1
)

# Save plot
pdf(
  "Output/Supp_Bilateral_Chord_SSP160_2050.pdf",
  width = 8,
  height = 8
)
circos.clear()
circos.par(start.degree = 90, gap.degree = 5)
chordDiagram(
  x = flow_matrix,
  grid.col = country_colors,
  transparency = 0.25,
  directional = 1,
  direction.type = c("arrows", "diffHeight"),
  link.arr.type = "big.arrow"
)
title(
  main = "Dominant Bilateral Migration Corridors (2050, SSP1–6.0)",
  cex.main = 1
)
dev.off()
```

##### END ######


