
This analysis investigates projected cross-border climate-driven migration and dengue risk across African countries under future socioeconomic and climate scenarios (SSPs) for 2020 and 2050. We quantify at-risk populations, compute absolute and percentage changes due to migration, and identify top countries and regions experiencing increases in risk. Spatial patterns are visualized through country- and region-level maps, bubble plots, and faceted regional maps, highlighting hotspots of exposure and change. Additionally, bilateral migration flows are mapped with curved arrows to illustrate directions and magnitudes, integrating environmental risk, population dynamics, and migration intensity. By combining temporal trends, country- and region-level aggregations, and multi-layered geospatial visualization, this workflow provides a comprehensive assessment of how climate-driven migration may alter dengue risk and population vulnerability across Africa.

Part 0: Libraries and setup
```{r}
# Load  Libraries
library(dplyr)
library(purrr)
library(tibble)
library(sf)
library(terra)
library(ggplot2)
library(ggrepel)
library(ggforce)
library(RColorBrewer)
library(patchwork)
library(scales)
library(viridis)

# Load migration flows
migration <- read.csv("Data/BilateralMigration/Summary_ACMI_bilateral_migration_2015_2050.csv")

# Load Africa shapefile with dengue risk
africa_sf <- read_sf("Data/Africa_regions.gpkg") %>%
  rename(denvr = denv_risk) %>%
  filter(!ISO3 %in% c("CPV", "STP", "SYC", "MUS", "REU", "COM"))  # remove small islands

# Convert to terra vector for raster operations
africa_vect <- vect(africa_sf)

# Simplified dataframe for joins (country ISO, name, region)
africa_df <- africa_sf %>%
  st_drop_geometry() %>%
  select(ISO2, ISO3, NAME, region)

# Load population raster stack (SSP projections)
pop_stack <- rast("Data/Population/Combined_Population_SSP1_SSP3_2020_2050.tif")

# Load dengue risk raster
denTP <- rast("Data/DTP_2020.tiff") %>%
  project("EPSG:4326")  # align CRS

# Align Africa vector CRS to raster
africa_vect <- project(africa_vect, "EPSG:4326")

# Create binary dengue risk mask
risk_threshold <- 0.5  
dengue_risk_binary <- classify(
  denTP,
  rcl = matrix(c(
    -Inf, risk_threshold, 0,  
    risk_threshold, Inf, 1    
  ), ncol = 3, byrow = TRUE)
)

migration_africa <- migration %>%
  filter(orig_iso3c %in% africa_df$ISO3 & dest_iso3c %in% africa_df$ISO3)

migration_joined <- migration_africa %>%
  left_join(africa_df %>% rename(orig_region = region, orig_name_full = NAME),
            by = c("orig_iso3c" = "ISO3")) %>%
  left_join(africa_df %>% rename(dest_region = region, dest_name_full = NAME),
            by = c("dest_iso3c" = "ISO3"))
```

Part 2: Compute Migration Inflows and Population at Risk
```{r}
#  Filter migration data to Africa only
migration_africa <- migration %>%
  filter(orig_iso3c %in% africa_df$ISO3 & dest_iso3c %in% africa_df$ISO3)

#  Join region and country info to migration data
migration_joined <- migration_africa %>%
  left_join(africa_df %>% rename(orig_region = region, orig_name_full = NAME),
            by = c("orig_iso3c" = "ISO3")) %>%
  left_join(africa_df %>% rename(dest_region = region, dest_name_full = NAME),
            by = c("dest_iso3c" = "ISO3"))

#  Define scenarios and corresponding columns
scenarios <- list(
  SSP126 = "GEPIC_WG2_SSP1_RCP26D_mean",  
  SSP160 = "GEPIC_WG2_SSP1_RCP60D_mean",  
  SSP360 = "GEPIC_WG2_SSP3_RCP60D_mean"   
)

# Prepare inflow per scenario
prepare_inflow <- function(df, scenario_name) {
  df %>%
    filter(year != 2015) %>%  # exclude base year
    group_by(dest_iso3c, year) %>%
    summarise(
      total_inflow = sum(abs(mean), na.rm = TRUE),
      .groups = "drop"
    ) %>%
    left_join(
      africa_df %>% select(ISO3, region, NAME),
      by = c("dest_iso3c" = "ISO3")
    ) %>%
    mutate(scenario = scenario_name)
}

#  Compute inflows for all scenarios
inflow_all <- purrr::map_dfr(names(scenarios), function(scenario) {
  prepare_inflow(
    migration_joined %>%
      select(dest_iso3c, year, mean = !!sym(scenarios[[scenario]])),
    scenario
  )
})

#  Final inflow table
inflow_final <- inflow_all %>%
  select(year, dest_iso3c, NAME, region, scenario, total_inflow) %>%
  arrange(region, NAME, year, scenario)

# Split inflows by scenario
migration_scenarios <- inflow_final %>%
  select(year, dest_iso3c, scenario, total_inflow) %>%
  rename(ISO3 = dest_iso3c) %>%
  group_by(scenario) %>%
  group_split(.keep = TRUE)

names(migration_scenarios) <- unique(inflow_final$scenario)

#  Load Population Raster Stack
pop_stack <- rast("Data/Population/Combined_Population_SSP1_SSP3_2020_2050.tif")

# Extract SSP and year from layer names
pop_layers_info <- data.frame(layer_name = names(pop_stack)) %>%
  mutate(
    SSP  = sub("_.*", "", layer_name),
    year = as.integer(sub(".*_", "", layer_name))
  )

# Map migration scenarios to SSPs
ssp_lookup <- c(
  SSP126 = "SSP1",
  SSP160 = "SSP1",
  SSP360 = "SSP3"
)

#  Load Dengue Risk Raster & Create Binary Mask
denTP <- rast("Data/DTP_2020.tiff")
target_crs <- "EPSG:4326"

denTP <- project(denTP, target_crs)
africa_vect <- project(africa_vect, target_crs)
africa_sf <- st_as_sf(africa_vect)

risk_threshold <- 0.5
dengue_risk_binary <- classify(
  denTP,
  rcl = matrix(c(-Inf, risk_threshold, 0,
                 risk_threshold, Inf, 1),
               ncol = 3, byrow = TRUE)
)

#  Compute Population at Risk with Migration
all_scenario_results <- list()

for (scenario_name in names(migration_scenarios)) {
  
  clim_migration <- migration_scenarios[[scenario_name]]
  ssp_scenario <- ssp_lookup[[scenario_name]]
  all_years_results <- list()
  
  for (year_i in unique(clim_migration$year)) {
    
    # --- Get corresponding population layer ---
    pop_layer_name <- pop_layers_info %>%
      filter(SSP == ssp_scenario & year == year_i) %>%
      pull(layer_name)
    
    if(length(pop_layer_name) == 0) next
    
    pop_layer <- pop_stack[[pop_layer_name]]
    
    # --- Aggregate & resample population to dengue raster resolution ---
    agg_factor <- round(res(denTP)[1] / res(pop_layer)[1])
    pop_agg <- aggregate(pop_layer, fact = agg_factor, fun = sum, na.rm = TRUE)
    pop_resampled <- resample(pop_agg, denTP, method = "bilinear")
    pop_masked <- mask(pop_resampled, africa_sf)
    
    # --- Population at risk ---
    pop_at_risk <- pop_masked * dengue_risk_binary
    
    # --- Baseline country-level population & risk ---
    baseline_risk <- data.frame(
      ISO3 = africa_sf$ISO3,
      country = africa_sf$NAME,
      region  = africa_sf$region,
      denvr   = africa_sf$denvr,
      total_pop_before_migration   = terra::extract(pop_masked, africa_vect, fun = sum, na.rm = TRUE)[,2],
      pop_at_risk_before_migration = terra::extract(pop_at_risk, africa_vect, fun = sum, na.rm = TRUE)[,2]
    ) %>%
      mutate(risk_proportion_before_migration = ifelse(total_pop_before_migration > 0,
                                                       pop_at_risk_before_migration / total_pop_before_migration, 0),
        year = year_i
      )
    
    # --- Join migration inflows ---
    migration_df <- clim_migration %>%
      filter(year == year_i) %>%
      select(ISO3, total_inflow)
    
    # --- Compute updated population & risk after migration ---
    risk_with_migration <- baseline_risk %>%
      left_join(migration_df, by = "ISO3") %>%
      mutate(
        total_inflow = if_else(is.na(total_inflow), 0, total_inflow),
        total_pop_after_migration   = total_pop_before_migration + total_inflow,
        pop_at_risk_after_migration = pop_at_risk_before_migration +
                                      (total_inflow * risk_proportion_before_migration),
        risk_proportion_after_migration = pop_at_risk_after_migration / total_pop_after_migration,
        scenario = scenario_name
      )
    
    all_years_results[[as.character(year_i)]] <- risk_with_migration
  }
  
  all_scenario_results[[scenario_name]] <- bind_rows(all_years_results)
}

# Combine all scenarios into one table
comparison_df_all_scenarios <- bind_rows(all_scenario_results) %>%
  mutate(
    risk_pop_change     = pop_at_risk_after_migration - pop_at_risk_before_migration,
    risk_pop_change_pct = ifelse(pop_at_risk_before_migration > 0,
                                 100 * risk_pop_change / pop_at_risk_before_migration, 0)
  )

summary_results <- comparison_df_all_scenarios %>%
  group_by(scenario, year) %>%
  summarise(
    total_pop_before_migration   = sum(total_pop_before_migration, na.rm = TRUE),
    total_pop_after_migration    = sum(total_pop_after_migration, na.rm = TRUE),
    pop_at_risk_before_migration = sum(pop_at_risk_before_migration, na.rm = TRUE),
    pop_at_risk_after_migration  = sum(pop_at_risk_after_migration, na.rm = TRUE),
    risk_pop_change              = sum(risk_pop_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    risk_pop_change_pct = 100 * (pop_at_risk_after_migration - pop_at_risk_before_migration) /
                          pop_at_risk_before_migration,
    risk_proportion_before = pop_at_risk_before_migration / total_pop_before_migration,
    risk_proportion_after  = pop_at_risk_after_migration / total_pop_after_migration
  ) %>%
  arrange(scenario, year)

# Population at Risk Change by Scenario
risk_scenario_plot <- ggplot(summary_results, aes(x = year, y = risk_pop_change, color = scenario)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  scale_y_continuous(labels = scales::comma) +
  scale_x_continuous(breaks = unique(summary_results$year)) +
  scale_color_brewer(palette = "Set2") +
  labs(
    y = "Population at Risk",
    x = "Year",
    color = "Scenario"
  ) +
  theme_bw(base_size = 14) +
  theme(
    axis.title = element_text(size = 20, face = "bold"),
    panel.grid = element_blank(),
    axis.text = element_text(size = 18),
    legend.title = element_text(size = 18, face = "bold"),
    legend.text = element_text(size = 16),
    legend.position = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA)
  )

# Print the plot
risk_scenario_plot

# Compute cumulative risk change by scenario over time
cumulative_results <- comparison_df_all_scenarios %>%
  arrange(scenario, year) %>%
  group_by(scenario) %>%
  mutate(cum_risk_pop_change = cumsum(risk_pop_change)) %>%
  ungroup()

# Summarize cumulative values by scenario and year
cumulative_summary <- cumulative_results %>%
  group_by(scenario, year) %>%
  summarise(
    cum_risk_pop_change = max(cum_risk_pop_change, na.rm = TRUE),
    .groups = "drop"
  )

# Cumulative Plot
risk_cumulative_plot <- ggplot(cumulative_summary, aes(
    x = year, y = cum_risk_pop_change / 1000,  # convert to thousands
    group = scenario, color = scenario
  )) +
  geom_line(size = 1.3) +
  geom_point(size = 2.5) +
  scale_y_continuous(labels = scales::comma) +  # labels still formatted nicely
  scale_x_continuous(breaks = seq(2020, 2050, by = 5)) +
  scale_color_manual(
    values = c(
      "SSP126" = "#008B00",
      "SSP160" = "#8B3626",
      "SSP360" = "#CD9B1D"
    ),
    labels = c("SSP1-2.6", "SSP1-6.0", "SSP3-6.0")
  ) +
  labs(
    y = "Population at Risk (Thousands)",  
    x = "Year",
    color = "Scenario"
  ) +
  theme_bw(base_size = 8) +
  theme(
    axis.title = element_text(size = 8, face = "bold"),
    panel.grid = element_blank(),
    axis.text = element_text(size = 8),
    legend.title = element_text(size = 8, face = "bold"),
    legend.text = element_text(size = 8),
    legend.position = c(0.05, 0.95),
    legend.justification = c("left", "top"),
    legend.background = element_rect(fill = "white", color = NA)
  )

risk_cumulative_plot

ggsave("Output/risk_cumulative_plot.pdf", 
       risk_cumulative_plot, 
       width = 5, 
       height = 3, 
       dpi = 300)
```

Part 3: Top 10 Most Affected Countries
```{r}
# Extract Top 10 Countries per Year
get_top10_per_year <- function(df, 
                               value_col, 
                               years = c(2020, 2050), 
                               n = 10, label = NULL) {
  df %>%
    filter(year %in% years, !is.na(ISO3)) %>%
    group_by(year) %>%
    slice_max(order_by = !!sym(value_col), n = n, with_ties = FALSE) %>%
    ungroup() %>%
    mutate(type = label)
}

# Reorder ISO3 factor levels by 2050 value
reorder_iso3_2050 <- function(df, value_col) {
  df_2050 <- df %>%
    filter(year == 2050, !is.na(ISO3)) %>%
    group_by(ISO3) %>%
    summarise(value_2050 = mean(!!sym(value_col), na.rm = TRUE)) %>%
    arrange(desc(value_2050))
  
  other_countries <- setdiff(unique(df$ISO3), df_2050$ISO3)
  new_levels <- c(df_2050$ISO3, other_countries)
  
  df %>%
    mutate(ISO3 = factor(ISO3, levels = new_levels))
}

# Plot Top 10 Bar Plot
plot_top10 <- function(df, value_col, y_label, title) {
  iso_codes <- levels(df$ISO3)
  color_palette <- colorRampPalette(brewer.pal(12, "Paired"))(length(iso_codes))
  
  ggplot(df, aes(x = factor(year), y = !!sym(value_col), fill = ISO3)) +
    geom_col(position = position_dodge(width = 0.8)) +
    scale_fill_manual(values = setNames(color_palette, iso_codes)) +
    labs(title = title, x = "Year", y = y_label, fill = "Country") +
    theme_bw(base_size = 14) +
    scale_y_continuous(breaks = pretty_breaks(n = 6)) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      axis.title = element_text(size = 16, face = "bold"),
      axis.text = element_text(size = 14),
      legend.title = element_text(size = 16, face = "bold"),
      legend.text = element_text(size = 14),
      legend.position = "right"
    )
}

# Prepare Top 10 Data for a Scenario
prepare_top10_data <- function(df_scenario) {
  
  # Top 10 by percentage increase
  top_pct <- get_top10_per_year(df_scenario, 
                                "risk_pop_change_pct", 
                                label = "Increase %")
  top_pct <- reorder_iso3_2050(top_pct, "risk_pop_change_pct")
  
  # Top 10 by total population at risk after migration (millions)
  top_after <- get_top10_per_year(df_scenario, 
                                  "pop_at_risk_after_migration", 
                                  label = "After Migration") %>%
    mutate(pop_at_risk_after_migration_mil = pop_at_risk_after_migration / 1e6)
  top_after <- reorder_iso3_2050(top_after, "pop_at_risk_after_migration_mil")
  
  # Top 10 by absolute increase (thousands)
  top_abs <- get_top10_per_year(df_scenario, "risk_pop_change", label = "Increase After Migration") %>%
    mutate(risk_pop_change_thousands = round(risk_pop_change / 1e3))
  top_abs <- reorder_iso3_2050(top_abs, "risk_pop_change_thousands")
  
  list(
    pct = top_pct,
    after = top_after,
    abs = top_abs
  )
}

# Generate Plots for a Scenario
plot_top10_scenario <- function(df_scenario, scenario_name) {
  top10_data <- prepare_top10_data(df_scenario)
  
  p1 <- plot_top10(top10_data$pct, 
                   "risk_pop_change_pct", 
                   "Percentage Increase", 
                   paste0(scenario_name, 
                          " – % Increase in At-Risk Population"))
  p2 <- plot_top10(top10_data$after, 
                   "pop_at_risk_after_migration_mil", 
                   "Population (millions)",
                   paste0(scenario_name, 
                          " – Population at Risk After Migration"))
  p3 <- plot_top10(top10_data$abs, 
                   "risk_pop_change_thousands", 
                   "Population (Thousands)",
                   paste0(scenario_name, 
                          " – Population Increase After Migration"))
  
  list(percent_increase = p1, total_after_mig = p2, absolute_increase = p3)
}

# SSP160 Only
ssp160_df <- comparison_df_all_scenarios %>%
  filter(scenario == "SSP160")

ssp160_plots <- plot_top10_scenario(ssp160_df, "SSP160")

# Display Plots
print(ssp160_plots$percent_increase)
print(ssp160_plots$total_after_mig)
print(ssp160_plots$absolute_increase)

# Export Plots
ggsave("Output/SSP160_Percent_Increase.png", 
       ssp160_plots$percent_increase, 
       width = 7, 
       height = 4.5, 
       dpi = 600)
ggsave("Output/SSP160_Total_AtRisk.pdf", 
       ssp160_plots$total_after_mig, 
       width = 7, 
       height = 4.5, 
       dpi = 600)
ggsave("Output/SSP160_Increase_AtRisk.pdf", 
       ssp160_plots$absolute_increase, 
       width = 7, 
       height = 4.5, 
       dpi = 600)

```

Part 4: Bubble Plot: At-Risk After Migration vs Absolute Change vs Percent Increase
```{r}
# Prepare Bubble Plot Data
prepare_bubble_data <- function(df_scenario, years = c(2020, 2050)) {
  df_scenario %>%
    filter(year %in% years, !is.na(ISO3)) %>%
    mutate(
      pop_at_risk_after_mil = pop_at_risk_after_migration / 1e6,
      risk_pop_change_thousands = risk_pop_change / 1e3
    )
}

# Plot Bubble for a Single Year
plot_bubble_year <- function(df_year) {
  iso_codes <- unique(df_year$ISO3)
  color_palette <- colorRampPalette(brewer.pal(12, "Paired"))(length(iso_codes))
  
  ggplot(df_year, aes(
    x = risk_pop_change_thousands,
    y = pop_at_risk_after_mil,
    color = ISO3,
    size = risk_pop_change_pct
  )) +
    geom_point(alpha = 0.8) +
    geom_text_repel(aes(label = ISO3), size = 3, show.legend = FALSE) +
    scale_y_log10(labels = comma) +  # Log scale for clarity
    scale_size_continuous(range = c(2, 10), name = "% Increase") +
    scale_color_manual(values = setNames(color_palette, iso_codes)) +
    labs(
      x = "Absolute Population Change (Thousands)",
      y = "Population at Risk After Migration (Millions, log scale)"
    ) +
    guides(color = "none") +
    theme_bw(base_size = 10) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      axis.title = element_text(size = 10, face = "bold"),
      axis.text = element_text(size = 9),
      legend.title = element_text(size = 9, face = "bold"),
      legend.text = element_text(size = 8),
      legend.position = c(0.85, 0.35)
    )
}

# Generate Bubble Plots for All Years
plot_bubble_scenario <- function(df_scenario, scenario_name, years = c(2020, 2050)) {
  bubble_data <- prepare_bubble_data(df_scenario, years)
  
  plots <- lapply(years, function(yr) {
    df_year <- bubble_data %>% filter(year == yr)
    plot <- plot_bubble_year(df_year) +
      ggtitle(paste0(scenario_name, " – Year ", yr))
    return(plot)
  })
  
  names(plots) <- paste0("Year_", years)
  return(plots)
}

# SSP160
ssp160_df <- comparison_df_all_scenarios %>% filter(scenario == "SSP160")
ssp160_bubble_plots <- plot_bubble_scenario(ssp160_df, "SSP160")

# Display Plots
print(ssp160_bubble_plots$Year_2020)
print(ssp160_bubble_plots$Year_2050)

# Save Plots
ggsave("Output/bubble_2020_SSP160.pdf", 
       ssp160_bubble_plots$Year_2020,
       width = 5, 
       height = 3, 
       dpi = 300)
ggsave("Output/bubble_2050_SSP160.pdf", 
       ssp160_bubble_plots$Year_2050, 
       width = 5, 
       height = 3, 
       dpi = 300)

```

Part 5: Change in Population at Risk Between 2050 and 2020
```{r}
# Prepare Top N Countries for Risk Change Plot
prepare_risk_change_data <- function(df_scenario, top_n = 15) {
  df_scenario %>%
    group_by(ISO3, country) %>%
    summarise(
      cumulative_risk = sum(risk_pop_change, na.rm = TRUE),
      risk_2020       = sum(risk_pop_change[year == 2020], na.rm = TRUE),
      risk_2050       = sum(risk_pop_change[year == 2050], na.rm = TRUE),
      risk_change     = risk_2050 - risk_2020,
      .groups = "drop"
    ) %>%
    mutate(
      cumulative_risk = as.integer(round(cumulative_risk, 0)),
      risk_2020       = as.integer(round(risk_2020, 0)),
      risk_2050       = as.integer(round(risk_2050, 0)),
      risk_change     = as.integer(round(risk_change, 0))
    ) %>%
    slice_max(order_by = cumulative_risk, n = top_n) %>%
    arrange(desc(cumulative_risk)) %>%
    mutate(
      country_factor = factor(ISO3, levels = rev(unique(ISO3[order(-cumulative_risk)])))
    )
}

# Plot Risk Change for Scenario
plot_risk_change <- function(df_top, scenario_name) {
  df_top <- df_top %>% mutate(change_positive = risk_change > 0)
  
  max_x <- max(df_top$cumulative_risk + df_top$risk_change, na.rm = TRUE)
  
  ggplot(df_top, aes(y = country_factor, x = cumulative_risk)) +
    geom_col(fill = "#CDCD00", alpha = 0.8) +
    geom_segment(aes(
      x = cumulative_risk,
      xend = cumulative_risk + risk_change,
      y = country_factor,
      yend = country_factor,
      color = change_positive
    ), size = 2) +
    geom_text(aes(
      x = cumulative_risk + risk_change,
      label = scales::comma(risk_change)
    ),
    hjust = ifelse(df_top$risk_change >= 0, -0.2, -0.6),
    color = "black",
    size = 3
    ) +
    scale_color_manual(
      values = c("TRUE" = "#EE0000", "FALSE" = "#008B45"),
      labels = c("TRUE" = "Increase from 2020 to 2050", "FALSE" = "Decrease from 2020 to 2050"),
      name = "Risk change"
    ) +
    scale_x_continuous(limits = c(0, max_x * 1.15), expand = expansion(mult = c(0, 0))) +
    labs(
      x = "Population at risk due to migration (2020–2050)",
      y = "Country",
      title = paste0(scenario_name, " – Change in At-Risk Population (Top ", nrow(df_top), " countries)")
    ) +
    theme_bw(base_size = 8) +
    theme(
      plot.background = element_rect(fill = "white", color = NA),
      panel.background = element_rect(fill = "white", color = NA),
      panel.grid = element_blank(),
      axis.title = element_text(size = 8, face = "bold"),
      axis.text.y = element_text(size = 8),
      legend.background = element_rect(fill = "white", color = NA),
      legend.title = element_text(size = 8, face = "bold"),
      legend.text = element_text(size = 8),
      legend.position = c(0.65, 0.25)
    )
}

# SSP160
ssp160_df <- comparison_df_all_scenarios %>% filter(scenario == "SSP160")
ssp160_risk_top <- prepare_risk_change_data(ssp160_df, top_n = 15)
p_ssp160_risk <- plot_risk_change(ssp160_risk_top, "SSP160")

# Display plot
print(p_ssp160_risk)

# Save plot
ggsave("Output/bar_plot_risk_SSP160.pdf",
       p_ssp160_risk,
       width = 4.25,
       height = 2.75,
       dpi = 300)

```

Part 6: Bubbles for Spatial Distribution of Population at Risk & Migration Change
```{r}
# Load Africa polygons 
africa <- read_sf("Data/Africa_regions.gpkg") %>%
  filter(!ISO3 %in% c("CPV", "STP", "SYC", "MUS", "REU", "COM"))

# Prepare spatial data with migration info
prepare_spatial_data <- function(migration_df, africa_sf) {
  # Join migration data to spatial polygons
  africa_sf %>%
    left_join(migration_df, by = "ISO3") %>%
    mutate(
      pop_at_risk_mil = pop_at_risk_after_migration / 1e6,
      denvr = factor(
        denvr,
        levels = c("Continuous risk", "Sporadic risk", "No Evidence")
      )
    )
}


# Generate centroid table for bubbles and labels
get_map_centroids <- function(spatial_df) {
  st_centroid(spatial_df) %>%
    mutate(
      X = st_coordinates(.)[,1],
      Y = st_coordinates(.)[,2]
    )
}

# Plot Africa map with bubbles
plot_africa_bubbles <- function(spatial_df, scenario_name) {
  centroids <- get_map_centroids(spatial_df)
  
  color_palette <- c(
    "Continuous risk" = "#8B2323",
    "Sporadic risk"   = "#FF7F24",
    "No Evidence"     = "#FFFACD"
  )
  
  ggplot() +
    # Dengue risk shading
    geom_sf(
      data = spatial_df,
      aes(fill = denvr),
      color = "#363636",
      size = 0.1,
      alpha = 0.5
    ) +
    
    # Migration bubbles (size scaled by absolute risk)
    geom_sf(
      data = centroids,
      aes(size = risk_pop_change),  
      color = "#27408B",
      alpha = 0.8,
      shape = 21,
      fill  = "#36648B", 
      stroke = 0.8
    ) +
    
    # Country labels with repulsion
    geom_text_repel(
      data = centroids,
      aes(x = X, y = Y, label = ISO3),
      size = 2,
      color = "#222222",
      box.padding = 0.4,
      point.padding = 0.5,
      force = 6,
      max.overlaps = Inf,
      segment.size = 0.2,
      min.segment.length = 0
    ) +
    
    # Facet by year
    facet_wrap(~year, ncol = 2, scales = "fixed") +
    
    # Fill and size scales
    scale_fill_manual(
      name   = "Dengue risk level",
      values = color_palette,
      na.value = "grey80"
    ) +
    scale_size_continuous(
      name   = "Population at risk (Thousands)",
      range  = c(1, 13),
      breaks = c(100, 1000, 2500, 5000, 10000),
      trans  = "sqrt"
    ) +
    
    # Map coordinates & minimal theme
    coord_sf() +
    theme_void(base_size = 10) +
    theme(
      strip.text = element_text(face = "bold", size = 10, color = "#333333"),
      legend.position = "right",
      legend.title = element_text(size = 10, face = "bold"),
      legend.text  = element_text(size = 10),
      plot.margin  = margin(10, 10, 10, 10)
    ) +
    labs(title = paste0(scenario_name, " – Spatial Distribution of Population at Risk"))
}


# SSP160
ssp160_spatial <- prepare_spatial_data(ssp160_filtered, africa)
population_map_ssp160 <- plot_africa_bubbles(ssp160_spatial, "SSP160")

# Display the map
print(population_map_ssp160)

# Save 
output_dir <- "Output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  filename = file.path(output_dir, 
                       "SSP160_Population_at_Risk_Map_2020_2050.pdf"),
  plot     = population_map_ssp160,
  width    = 8,
  height   = 5,
  dpi      = 300
)

```

Part 7:  Regional Map of Population at Risk & Migration Change 
```{r}
# Parameters
selected_years <- c(2020, 2050)
selected_scenario <- "SSP160"

# Prepare regional data
regional_data <- comparison_df_all_scenarios %>%
  filter(year %in% selected_years,
         scenario == selected_scenario) %>%
  left_join(africa %>% st_as_sf(), by = "ISO3") %>%
  st_as_sf() %>%
  group_by(region.x, year) %>%
  summarise(
    pop_at_risk_after_migration = sum(pop_at_risk_after_migration, na.rm = TRUE),
    risk_pop_change = sum(risk_pop_change, na.rm = TRUE),
    geometry = st_union(geom),
    .groups = "drop"
  ) %>%
  mutate(
    pop_at_risk_mil = pop_at_risk_after_migration / 1e6,
    risk_pop_change_label = scales::comma(round(risk_pop_change, 0))
  )

# Compute centroids for circles & labels
region_centroids <- st_centroid(regional_data) %>%
  mutate(
    X = st_coordinates(.)[,1],
    Y = st_coordinates(.)[,2]
  )

# Plot function
plot_regional_map <- function(regional_sf, centroids_sf, title = NULL) {
  ggplot() +
    
    # Fill by population at risk
    geom_sf(
      data = regional_sf,
      aes(fill = pop_at_risk_mil),
      color = "#363636",
      size = 0.5,
      alpha = 0.8
    ) +
    
    # Africa outline
    geom_sf(
      data = africa %>% st_as_sf(),
      fill = NA,
      color = "#5C5C5C",
      size = 0.2
    ) +
    
    # Circles at centroid scaled by population at risk
    geom_sf(
      data = centroids_sf,
      aes(size = pop_at_risk_mil),
      color = "#0F0F0F",
      fill  = "#0F0F0F",
      shape = 21,
      alpha = 0.7,
      show.legend = FALSE
    ) +
    
    scale_size_continuous(range = c(17, 30)) +
    
    # Labels for change in population at risk
    geom_sf_text(
      data = centroids_sf,
      aes(label = risk_pop_change_label),
      size = 5,
      color = "white",
      fontface = "bold"
    ) +
    
    # Color gradient for population at risk
    scale_fill_gradient(
      name = "Population at \nRisk (Millions)",
      low  = "#F0F0F0",
      high = "#EE2C2C",
      na.value = "lightgrey"
    ) +
    
    # Facet by year
    facet_wrap(~year) +
    
    # Coordinate limits
    coord_sf(xlim = c(-20, 55), ylim = c(-40, 40), expand = FALSE) +
    
    # Minimal theme
    theme_void() +
    theme(
      legend.position = "right",
      legend.title = element_text(size = 12, face = "bold"),
      legend.text  = element_text(size = 12),
      strip.text   = element_text(size = 12, face = "bold"),
      plot.title   = element_text(hjust = 0.5, size = 14, face = "bold")
    ) +
    labs(title = title)
}

# Generate the regional map
regional_map_ssp160 <- plot_regional_map(regional_data, region_centroids, 
                                         title = "SSP160 – Population at Risk & Change by Region")

# Display the map
print(regional_map_ssp160)

# Export the faceted map
output_dir <- "Output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  filename = file.path(output_dir, 
                       "Regional_Map_SSP160_2020_2050.pdf"),
  plot = regional_map_ssp160,
  width = 10,
  height = 5,
  dpi = 600
)
```

Part 8: Bar Plot of Absolute Population Change at Risk by Region  
```{r}
# Parameters
selected_years <- c(2020, 2050)
selected_scenario <- "SSP160"

# Aggregate total population at risk by region and year
regional_risk <- comparison_df_all_scenarios %>%
  filter(year %in% selected_years, scenario == selected_scenario) %>%
  group_by(region, year) %>%
  summarise(total_risk = sum(risk_pop_change, na.rm = TRUE), .groups = "drop") %>%
  mutate(total_risk_k = total_risk / 1000)  

# Function to create regional bar plot for a single year
plot_regional_bar <- function(data, plot_year) {
  
  data_year <- data %>% filter(year == plot_year)
  
  ggplot(data_year, aes(x = reorder(region, total_risk_k), y = total_risk_k, fill = region)) +
    geom_col(alpha = 0.85, color = "black") +
    scale_fill_brewer(palette = "Set2", name = "Region") +
    scale_y_continuous(labels = scales::comma) +
    labs(
      y = "Population at Risk (Thousands)",
      x = NULL,
      title = paste0("Regional Population at Risk – ", plot_year)
    ) +
    theme_bw(base_size = 10) +
    theme(
      axis.text.x        = element_text(angle = 30, hjust = 1, face = "bold"),  
      axis.text.y        = element_text(face = "bold"),
      strip.background   = element_blank(),
      panel.grid         = element_blank(),
      panel.background   = element_blank(),
      legend.position    = "none",
      plot.title         = element_text(size = 12, face = "bold", hjust = 0.5)
    )
}

# Generate plots for each year
regional_bar_2020 <- plot_regional_bar(regional_risk, 2020)
regional_bar_2050 <- plot_regional_bar(regional_risk, 2050)

# Display plots
print(regional_bar_2020)
print(regional_bar_2050)

# Save 
output_dir <- "Output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  filename = file.path(output_dir, 
                       "regional_risk_bar_2020.pdf"),
  plot     = regional_bar_2020,
  width    = 4,
  height   = 3,
  dpi      = 300
)

ggsave(
  filename = file.path(output_dir, 
                       "regional_risk_bar_2050.pdf"),
  plot     = regional_bar_2050,
  width    = 4,
  height   = 3,
  dpi      = 300
)
```

Part 9: Dengue Risk and Migration Flow Directions 
```{r}
# Load Africa shapefile and filter islands
africa <- read_sf("Data/Africa_regions.gpkg") %>%
  filter(!ISO3 %in% c("CPV", "STP", "SYC", "MUS", "REU", "COM"))

# Order Dengue Risk Levels
africa <- africa %>%
  mutate(
    denv_risk = factor(
      denv_risk,
      levels = c("Continuous risk", 
                 "Sporadic risk", 
                 "No Evidence")
    )
  )

# Compute country centroids for plotting
country_centroids <- africa %>%
  filter(ISO3 != "NAM") %>%
  st_centroid() %>%
  dplyr::select(ISO3, NAME, geom) %>%
  mutate(
    lon = st_coordinates(geom)[, 1],
    lat = st_coordinates(geom)[, 2]
  ) %>%
  st_drop_geometry()

# Load bilateral migration data
migration <- read.csv("Data/BilateralMigration/Summary_ACMI_bilateral_migration_2015_2050.csv")

# Select SSP column and year
ssp_col <- "GEPIC_WG2_SSP1_RCP60D_mean"
plot_year <- 2050

# Prepare Top Migration Flows
top_flows <- migration %>%
  filter(!orig_iso3c %in% "NAM", !dest_iso3c %in% "NAM") %>%
  select(orig_iso3c, dest_iso3c, year, mean = all_of(ssp_col)) %>%
  filter(year == plot_year, !is.na(orig_iso3c), !is.na(dest_iso3c)) %>%
  slice_max(order_by = mean, n = 150) %>%
  # Join origin coordinates
  left_join(country_centroids, by = c("orig_iso3c" = "ISO3")) %>%
  rename(orig_lon = lon, orig_lat = lat, orig_name = NAME) %>%
  # Join destination coordinates
  left_join(country_centroids, by = c("dest_iso3c" = "ISO3")) %>%
  rename(dest_lon = lon, dest_lat = lat, dest_name = NAME) %>%
  mutate(flow_id = row_number())

# Function to generate Bezier curves
generate_bezier <- function(df) {
  dx <- df$dest_lon - df$orig_lon
  dy <- df$dest_lat - df$orig_lat
  
  mid_x <- (df$orig_lon + df$dest_lon) / 2
  mid_y <- (df$orig_lat + df$dest_lat) / 2
  
  length <- sqrt(dx^2 + dy^2)
  perp_x <- -dy / length
  perp_y <- dx / length
  
  curve_h <- 5  # adjust curve height
  control_x <- mid_x + perp_x * curve_h
  control_y <- mid_y + perp_y * curve_h
  
  tibble(
    flow_id = df$flow_id,
    x = c(df$orig_lon, control_x, df$dest_lon),
    y = c(df$orig_lat, control_y, df$dest_lat),
    flow_size = df$mean
  )
}

# Compute Bezier paths and log-scale flow size
bezier_paths <- top_flows %>%
  group_by(flow_id) %>%
  group_modify(~ generate_bezier(.x)) %>%
  ungroup() %>%
  mutate(flow_size_log = log1p(flow_size))

# Plot Dengue Risk + Migration Flows
plot_migration_flows <- ggplot() +
  
  # Dengue risk background
  geom_sf(
    data = africa,
    aes(fill = denv_risk),
    color = "#636363",
    linewidth = 0.2,
    alpha = 0.6
  ) +
  
  # Country labels
  geom_text_repel(
    data = country_centroids,
    aes(x = lon, y = lat, label = ISO3),
    size = 2,
    color = "black",
    max.overlaps = Inf,
    box.padding = 0.3,
    point.padding = 0.2,
    segment.color = "grey60",
    segment.size = 0.2
  ) +
  
  # Migration flows (Bezier)
  geom_bezier(
    data = bezier_paths,
    aes(
      x = x,
      y = y,
      group = flow_id,
      color = flow_size_log,
      size = flow_size_log
    ),
    lineend = "round",
    arrow = arrow(length = unit(0.12, "inches"), type = "closed", angle = 25)
  ) +
  
  # Flow size scaling
  scale_size(range = c(0.05, 1), name = "Flow size (log)") +
  
  # Country points
  geom_point(
    data = country_centroids,
    aes(x = lon, y = lat),
    shape = 21,
    fill = "#0F0F0F",
    color = "#0F0F0F",
    size = 1,
    alpha = 0.8
  ) +
  
  # Dengue risk fill
  scale_fill_manual(
    values = c(
      "Continuous risk" = "#8B2323",
      "Sporadic risk"   = "#FF7F24",
      "No Evidence"     = "#FFFACD"
    ),
    name = "Dengue Risk Level"
  ) +
  
  # Flow color gradient
  scale_color_gradientn(
    colours = colorRampPalette(c("#1874CD", "#EEEE00", "#FFA54F", "#FF0000"))(100),
    name = "Flow (log scale)",
    labels = function(x) scales::comma(round(expm1(x))),
    guide = guide_colorbar(barwidth = 1.2, barheight = 10)
  ) +
  
  theme_void(base_size = 14) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 12),
    plot.margin = unit(rep(0.4, 4), "cm")
  )


# Display plot
print(plot_migration_flows)

# Save 
output_dir <- "Output"
if (!dir.exists(output_dir)) dir.create(output_dir, recursive = TRUE)

ggsave(
  filename = file.path(output_dir, "Dengue_Migration_Flows_2050.pdf"),
  plot = plot_migration_flows,
  width = 6,
  height = 4,
  dpi = 300
)
```







