
This analysis investigates projected *internal migration* flows across African countries under future climate and socioeconomic scenarios (SSPs) for 2020 and 2050. We quantify total inflows and outflows at the country level, identify top origin and destination areas, map spatial distributions of migration intensity, and highlight hotspot regions with high likelihoods of departure or arrival. By integrating temporal trends, country-level aggregations, and spatial visualization, the work provides a comprehensive view of internal climate-driven migration patterns.

Part 0 - Libraries & Setup
```{r}
# Load required libraries
library(dplyr)
library(tidyr)
library(sf)
library(terra)
library(stringr)
library(ggplot2)
library(scales)
library(ggnewscale)
library(ggspatial)

# Load Africa boundaries
africa_sf <- st_read("Data/Africa_regions.gpkg", quiet = TRUE)
```

Part 1 - Internal Migration Data Preparation
```{r}
# Load internal migration data
migration_sf <- st_read("Data/InternalMigration/combined_internal_migration.gpkg", quiet = TRUE)

# Keep only columns ending with "_diff"
flow_data <- migration_sf %>% select(ends_with("_diff"))

# Convert from wide to long format
flow_long <- flow_data %>%
  pivot_longer(
    cols = ends_with("_diff"),
    names_to = "variable",
    values_to = "migration_flow"
  ) %>%
  mutate(
    # Extract RCP, SSP, and year
    RCP_num = str_sub(variable, 3, 4),
    SSP_num = str_sub(variable, 5, 5),
    year = paste0("20", str_sub(variable, 6, 7)),
    scenario = paste0("SSP", SSP_num, RCP_num),
    flow_class = case_when(
      migration_flow > 0 ~ "Inflow",
      migration_flow < 0 ~ "Outflow",
      TRUE ~ "Neutral"
    )
  ) %>%
  select(-RCP_num, -SSP_num)
```

Part 2 - Total Inflow and Outflow Time Series by Scenario
```{r}
# Aggregate by country, scenario, and flow class
country_flows_sf <- flow_long %>%
  st_join(africa_sf %>% select(ISO3)) %>%
  st_set_geometry(NULL) %>%
  group_by(ISO3, scenario, year, flow_class) %>%
  summarise(total_migration_flow = sum(migration_flow, na.rm = TRUE), .groups = "drop") %>%
  left_join(africa_sf %>% select(ISO3, geometry), by = "ISO3") %>%
  st_as_sf() %>%
  pivot_wider(
    names_from  = flow_class,
    values_from = total_migration_flow,
    values_fill = 0
  )

# Summarize total inflows by scenario and year
inflow_flows <- country_flows_sf %>%
  st_set_geometry(NULL) %>%
  group_by(scenario, year) %>%
  summarise(total_inflow = sum(Inflow, na.rm = TRUE), .groups = "drop") %>%
  filter(scenario %in% c("SSP126", "SSP160", "SSP360")) %>%
  mutate(
    scenario = factor(scenario, levels = c("SSP126", "SSP160", "SSP360"),
                      labels = c("SSP1-2.6", "SSP1-6.0", "SSP3-6.0")),
    year = as.numeric(year)
  )

# Plot total inflows
inflow_plot <- ggplot(inflow_flows, aes(x = year, y = total_inflow / 1e6, color = scenario)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = unique(inflow_flows$year)) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_brewer(palette = "Set2") +
  coord_cartesian(ylim = c(14, 80)) +
  labs(x = "Year", y = "Migrants (millions)", color = "Scenario") +
  theme_bw(base_size = 12, base_family = "Helvetica") +
  theme(
    panel.grid       = element_blank(),
    axis.title       = element_text(face = "bold"),
    legend.title     = element_text(face = "bold"),
    legend.position  = c(0.35, 0.45),
    legend.justification = c("right", "bottom"),
    legend.background = element_rect(fill = alpha("white", 0.7), color = NA)
  )

inflow_plot
ggsave("Output/Inflow_by_scenario_Millions.pdf", plot = inflow_plot, width = 4, height = 3, dpi = 300)
```

Part 3 - Top 10 Countries (Inflows & Outflows)
```{r}
# Filter only SSP160
ssp160_data <- country_flows_sf %>% filter(scenario == "SSP160")

# Identify top 10 origins (Outflows) and destinations (Inflows)
top10_origins      <- ssp160_data %>% st_set_geometry(NULL) %>%
  group_by(ISO3) %>% summarise(total_outflow = sum(abs(Outflow), na.rm = TRUE)) %>%
  arrange(desc(total_outflow)) %>% slice_head(n = 10) %>% pull(ISO3)

top10_destinations <- ssp160_data %>% st_set_geometry(NULL) %>%
  group_by(ISO3) %>% summarise(total_inflow = sum(Inflow, na.rm = TRUE)) %>%
  arrange(desc(total_inflow)) %>% slice_head(n = 10) %>% pull(ISO3)

# Prepare data for bar plots
outflow_data <- ssp160_data %>%
  st_set_geometry(NULL) %>%
  filter(ISO3 %in% top10_origins) %>%
  group_by(ISO3, year) %>%
  summarise(yearly_outflow = sum(abs(Outflow), na.rm = TRUE), .groups = "drop") %>%
  mutate(ISO3 = factor(ISO3, levels = top10_origins)) %>%
  filter(year %in% c("2020", "2030", "2040", "2050"))

inflow_data <- ssp160_data %>%
  st_set_geometry(NULL) %>%
  filter(ISO3 %in% top10_destinations) %>%
  group_by(ISO3, year) %>%
  summarise(yearly_inflow = sum(Inflow, na.rm = TRUE), .groups = "drop") %>%
  mutate(ISO3 = factor(ISO3, levels = top10_destinations)) %>%
  filter(year %in% c("2020", "2030", "2040", "2050"))

# Plotting function
plot_migration <- function(data, yvar, fillvar, y_label = "") {
  ggplot(data, aes(x = factor(year), y = !!sym(yvar)/1e6, fill = !!sym(fillvar))) +
    geom_col(position = "dodge", color = "black") +
    scale_y_continuous(labels = scales::comma) +
    scale_fill_brewer(palette = "Paired") +
    labs(x = "Year", y = y_label, fill = "Country") +
    theme_bw(base_size = 12, base_family = "Helvetica") +
    theme(
      axis.title = element_text(size = 10, face = "bold"),
      axis.text  = element_text(size = 10),
      legend.title = element_text(size = 10, face = "bold"),
      legend.text  = element_text(size = 10),
      legend.position = "right"
    )
}

# Generate plots
outflow_plot <- plot_migration(outflow_data, 
                               "yearly_outflow", 
                               "ISO3", 
                               "Outflow (Millions)")
inflow_plot  <- plot_migration(inflow_data,  
                               "yearly_inflow",  
                               "ISO3", 
                               "Inflow (Millions)")

outflow_plot
inflow_plot

# Save
ggsave("Output/Top10_Outflow_Countries_2020_2050.pdf", 
       outflow_plot, 
       width = 4, 
       height = 3, 
       dpi = 300)
ggsave("Output/Top10_Inflow_Countries_2020_2050.pdf",  
       inflow_plot,  
       width = 4, 
       height = 3, 
       dpi = 300)

```

Part 4 - Spatial Distribution of Internal Migration (2020 & 2050)
```{r}
# Filter SSP160 and target years
target_years <- c("2020", "2050")
flow_filtered <- flow_long %>% filter(scenario == "SSP160", year %in% target_years)
flow_filtered$year <- factor(flow_filtered$year, levels = target_years)

# Split into Outflow and Inflow
leaving  <- flow_filtered %>% filter(flow_class == "Outflow") %>% mutate(migration_flow = abs(migration_flow))
arriving <- flow_filtered %>% filter(flow_class == "Inflow")

# Bin flows
breaks <- c(0, 100, 1e3, 1e4, 1e5, 1e6, Inf)
labels <- c("0", "100", "1k", "10k", "100k", "1M+")
leaving  <- leaving %>% mutate(flow_bin = cut(migration_flow, breaks = breaks, labels = labels, include.lowest = TRUE))
arriving <- arriving %>% mutate(flow_bin = cut(migration_flow, breaks = breaks, labels = labels, include.lowest = TRUE))

# Define colors
red_colors   <- c("0"="#ffe5e0","100"="#ff9990","1k"="#ff6666","10k"="#ff1a1a","100k"="#800000","1M+"="#4d0000")
green_colors <- c("0"="#e0ffe0","100"="#99ff99","1k"="#66ff66","10k"="#1aff1a","100k"="#009900","1M+"="#004d00")

# Plot map
migration_map <- ggplot() +
  geom_sf(data = leaving, aes(fill = flow_bin, geometry = geom), color = NA) +
  scale_fill_manual(name = "Outflow", values = red_colors) +
  ggnewscale::new_scale_fill() +
  geom_sf(data = arriving, aes(fill = flow_bin, geometry = geom), color = NA) +
  scale_fill_manual(name = "Inflow", values = green_colors) +
  geom_sf(data = africa_sf, fill = NA, color = "#303030", size = 0.5) +
  geom_sf_text(data = africa_sf, aes(label = ISO3), size = 3, color = "#474747") +
  facet_wrap(~year, ncol = 2) +
  theme_void() +
  theme(legend.position = "bottom", legend.title = element_text(face="bold", size=10)) +
  coord_sf(xlim = c(-20, 55), ylim = c(-40, 40), expand = FALSE)

migration_map
ggsave("Output/imssp160.pdf", 
       migration_map, 
       width = 10, 
       height = 6, 
       dpi = 600)
```

Part 5 - Migration Hotspots (Origin vs Destination)
```{r}
hotspot_sf <- migration_sf %>% select(Hotspot20, Hotspot50, geom)
hotspot_long <- hotspot_sf %>%
  pivot_longer(cols = starts_with("Hotspot"), names_to = "year", values_to = "hotspot_value") %>%
  mutate(
    year = paste0("20", gsub("Hotspot", "", year)),
    type = case_when(hotspot_value > 0 ~ "Destination",
                     hotspot_value < 0 ~ "Origin",
                     TRUE ~ "Neutral"),
    class = case_when(hotspot_value == 4 ~ "Very Likely",
                      hotspot_value == 2 ~ "Likely",
                      hotspot_value == -2 ~ "Likely",
                      hotspot_value == -4 ~ "Very Likely",
                      TRUE ~ "Neutral")
  ) %>%
  filter(type != "Neutral") %>%
  mutate(class_factor = factor(class, levels = c("Likely","Very Likely")))

origin_sf      <- hotspot_long %>% filter(type=="Origin")
destination_sf <- hotspot_long %>% filter(type=="Destination")

hotspot <- ggplot() +
  geom_sf(data = origin_sf, aes(fill = class_factor), color = NA) +
  scale_fill_manual(name="Origin", values=c("Likely"="#ff1a1a","Very Likely"="#4d0000")) +
  ggnewscale::new_scale_fill() +
  geom_sf(data = destination_sf, aes(fill = class_factor), color = NA) +
  scale_fill_manual(name="Destination", values=c("Likely"="#1aff1a","Very Likely"="#004d00")) +
  geom_sf(data = africa_sf, fill = NA, color="#303030", size=0.2) +
  facet_wrap(~year, ncol=2) +
  theme_void(base_size=10) +
  theme(strip.text=element_text(face="bold",size=10), legend.position="bottom",
        legend.title=element_text(face="bold",size=10), legend.text=element_text(size=10)) +
  coord_sf(xlim=c(-20,55), ylim=c(-40,40), expand=FALSE)

hotspot
ggsave("Output/hotspot.pdf", 
       hotspot, 
       width=10, 
       height=6, 
       dpi=300)
```

