
# Climate-Driven Internal Migration and Dengue Risk in Africa

This analysis investigates projected climate-driven internal migration across African countries and its potential impact on population exposure to dengue risk under future socioeconomic and climate scenarios (SSPs) for 2020 and 2050. The approach integrates population-at-risk data with dengue suitability maps and internal migration projections. Population inflows are added to baseline populations to compute risk with and without migration. Absolute and percentage changes are calculated at grid, country, and regional levels, with spatial and temporal visualization to identify exposure hotspots.


Part 0 – Load Libraries and Datasets
```{r}
# ==============================
# Load Required Libraries
# ==============================
library(dplyr)
library(tidyr)
library(sf)
library(terra)
library(stringr)
library(ggplot2)
library(scales)
library(RColorBrewer)
library(patchwork)
library(ggforce)
library(ggrepel)
library(tibble)
library(purrr)
library(viridis)

# Load Datasets
migration_orig <- st_read("Data/InternalMigration/combined_internal_migration.gpkg", quiet = TRUE)
denTP <- rast("Data/DTP_2020.tiff")
pop_stack <- rast("Data/Population/Population_2020_2050_ACMI.tif")
africa_sf <- st_as_sf(vect("Data/Africa_regions.gpkg"))
```

Part 1 - Spatial distribution of dengue risk 
```{r, warning=FALSE}
# Select columns with migration differences and reshape
flow_long <- migration_orig %>%
  st_drop_geometry() %>%
  select(ends_with("_diff")) %>%
  pivot_longer(
    cols = everything(),
    names_to = "variable",
    values_to = "migration_flow"
  ) %>%
  mutate(
    RCP_num = str_sub(variable, 3, 4),
    SSP_num = str_sub(variable, 5, 5),
    year = paste0("20", str_sub(variable, 6, 7)),
    scenario = paste0("SSP", SSP_num, RCP_num)
  ) %>%
  select(-RCP_num, -SSP_num)

# Binary dengue suitability raster
dengue_bin <- classify(denTP, rbind(c(-Inf, 0.5, 0), c(0.5, Inf, 1)))

# Aggregate population to ~10 km resolution
fact <- round(res(dengue_bin)[1] / res(pop_stack)[1])
pop_agg <- aggregate(pop_stack, fact = fact, fun = sum, na.rm = TRUE)

# Resample dengue raster to population grid
dengue_res <- resample(crop(dengue_bin, pop_agg[[1]]), pop_agg[[1]], method = "near")

#Compute Dengue Risk With and Without Migration
years_analysis <- c(2020, 2050)
scenarios <- "SSP160"
all_years_df <- list()

for (y in years_analysis) {
  pop_layer <- pop_agg[[grep(as.character(y), names(pop_agg))]]
  risk_no_mig <- mask(pop_layer, dengue_res, maskvalue = 0, updatevalue = 0)

  for (ssp in scenarios) {
    # Subset migration flows
    mig_sub <- flow_long %>% filter(year == as.character(y), scenario == ssp)
    
    if (nrow(mig_sub) == 0) {
      mig_rast <- rast(pop_layer); values(mig_rast) <- 0
    } else {
      mig_sf <- migration_orig %>%
        select(geom) %>%
        bind_cols(mig_sub %>% select(migration_flow))
      mig_rast <- rasterize(vect(mig_sf), pop_layer, field = "migration_flow", fun = sum, background = 0)
    }
    
    # Population with migration
    pop_with_mig <- pop_layer + mig_rast
    risk_with_mig <- mask(pop_with_mig, dengue_res, maskvalue = 0, updatevalue = 0)
    
    # Absolute and percentage change
    absolute_change <- risk_with_mig - risk_no_mig
    percentage_change <- ifel(risk_no_mig == 0, NA, (absolute_change / risk_no_mig) * 100)
    
    # Combine rasters and convert to dataframe
    raster_stack <- c(risk_no_mig, risk_with_mig, absolute_change, percentage_change)
    names(raster_stack) <- c("risk_without_migration", "risk_with_migration",
                             "absolute_change_dengue_risk", "percentage_change")
    
    df <- as.data.frame(raster_stack, xy = TRUE, na.rm = TRUE)
    df$year <- y; df$scenario <- ssp
    all_years_df[[paste(y, ssp, sep = "_")]] <- df
  }
}

combined_df <- bind_rows(all_years_df)

# Categorize and Visualize Dengue Risk
combined_df <- combined_df %>%
  mutate(
    pop_cat_with_mig = cut(risk_with_migration / 1000,
                           breaks = c(-Inf,1,5,10,50,150,Inf),
                           labels = c("0-1","1-5","5-10","10-50","50-150",">150"),
                           include.lowest = TRUE, right = FALSE),
    abs_change_cat = cut(absolute_change_dengue_risk,
                         breaks = c(-Inf, -1000, -500, -100, -1, 0, 1, 100, 500, 1000, Inf),  
                         labels = c("<-1000","-1000–-500","-500–-100","-100–-1","-1–0",
                                    "0–1","1–100","100–500","500–1000",">1000"),
                         include.lowest = TRUE, right = FALSE)
  )

# Color palettes
seq_colors <- c("#ffffcc", "#ffeda0", "#feb24c", "#f03b20", "#bd0026", "#800026")
div_colors <- c(
  "#053061","#2166ac","#4393c3","#92c5de","#d1e5f0",
  "#FFFFFF",
  "#fddbc7","#f4a582","#b2182b","#bd0026","#800026"
)

# Population at risk WITH migration
plot_with_mig <- ggplot(combined_df) +
  geom_tile(aes(x = x, y = y, fill = pop_cat_with_mig)) +
  geom_sf(data = africa_sf, fill = NA, color = "#303030", size = 0.5) +
  scale_fill_manual(values = seq_colors, name = "Population at risk (Thousands)",
                    drop = FALSE, guide = guide_legend(title.position="top", 
                                                       direction="horizontal", nrow=1)) +
  geom_sf_text(data = africa_sf, aes(label = ISO3), size = 2, color="#474747", check_overlap = TRUE) +
  facet_wrap(~year, ncol = 2) +
  theme_void() + theme(plot.title = element_text(hjust=0.5, size=14, face="bold"),
                       legend.position="bottom", legend.text=element_text(size=8),
                       legend.title=element_text(size=10, face="bold"),
                       strip.text=element_text(face="bold", size=12)) +
  coord_sf(xlim=c(-20,55), ylim=c(-40,40), expand=FALSE)

ggsave("Output/Population_With_Migration_All_SSP.pdf", 
       plot_with_mig, 
       width=8, 
       height=6, 
       dpi=300)

# Absolute change due to migration
plot_abs_change <- ggplot(combined_df) +
  geom_tile(aes(x = x, y = y, fill = abs_change_cat)) +
  geom_sf(data = africa_sf, fill = NA, color = "#303030", size = 0.5) +
  scale_fill_manual(values = div_colors, name="Risk change",
                    drop=FALSE, guide = guide_legend(title.position="top", 
                                                     direction="horizontal", nrow=1)) +
  facet_wrap(~year, ncol=2) +
  theme_void() + theme(legend.position="bottom", legend.text=element_text(size=8),
                       strip.text=element_text(face="bold", size=12)) +
  coord_sf(xlim=c(-20,55), ylim=c(-40,40), expand=FALSE)

ggsave("Output/Absolute_Change_Migration_All_SSP.pdf", 
       plot_abs_change, 
       width=8, 
       height=6, 
       dpi=300)
```

Part 2 - Country-Level Bar Plots
```{r}
# Convert to sf and join country boundaries
combined_sf <- st_as_sf(combined_df, coords = c("x","y"), crs=st_crs(africa_sf))
combined_sf <- st_join(combined_sf, africa_sf %>% select(ISO3)) %>% filter(!is.na(ISO3))

# Summarize absolute change per country
country_change <- combined_sf %>%
  st_drop_geometry() %>%
  group_by(ISO3, year, scenario) %>%
  summarize(
    increase = sum(absolute_change_dengue_risk[absolute_change_dengue_risk > 0], na.rm=TRUE),
    decrease = sum(absolute_change_dengue_risk[absolute_change_dengue_risk < 0], na.rm=TRUE),
    .groups="drop"
  ) %>%
  pivot_longer(cols=c(increase,decrease), names_to="Risk", values_to="value") %>%
  mutate(value_million = value / 1e6)

# Select top 20 countries across all scenarios
top_countries <- country_change %>%
  group_by(ISO3) %>% summarize(total_change=sum(abs(value_million)), .groups="drop") %>%
  arrange(desc(total_change)) %>% slice(1:20) %>% pull(ISO3)
plot_data_top <- country_change %>% filter(ISO3 %in% top_countries)

# Define plotting order
country_order <- plot_data_top %>% group_by(ISO3) %>% summarize(total_change=sum(abs(value_million))) %>%
  arrange(desc(total_change)) %>% pull(ISO3)

# Function to plot bar chart per scenario-year
make_plot <- function(df, scenario_name, year_value){
  df_year <- df %>% filter(scenario==scenario_name, year==year_value)
  ggplot(df_year, aes(x=factor(ISO3, levels=country_order), y=value_million, fill=Risk)) +
    geom_col() + coord_flip() +
    scale_fill_manual(values=c("increase"="firebrick","decrease"="steelblue"),
                      labels=c("Increase","Decrease")) +
    labs(title=paste0("Scenario: ", scenario_name, " — Year ", year_value),
         x="Country", y="Population at risk (Millions)") +
    theme_bw() +
    theme(axis.text.x=element_text(size=10, face="bold"),
          axis.title=element_text(size=10, face="bold"),
          legend.title=element_text(size=10, face="bold"),
          legend.text=element_text(size=10),
          legend.position=c(0.95,0.65),
          legend.justification=c("right","bottom"))
}

# Generate and save plots
scenario_list <- unique(plot_data_top$scenario)
plots_2020 <- lapply(scenario_list, function(sc) make_plot(plot_data_top, sc, 2020))
plots_2050 <- lapply(scenario_list, function(sc) make_plot(plot_data_top, sc, 2050))

dir.create("Output", showWarnings = FALSE)
for (i in seq_along(scenario_list)){
  ggsave(paste0("Output/change2020_", 
                scenario_list[i], ".pdf"), 
         plots_2020[[i]], 
         width=5, 
         height=6, 
         dpi=300)
  ggsave(paste0("Output/change2050_", 
                scenario_list[i], ".pdf"), 
         plots_2050[[i]], 
         width=5, 
         height=6, 
         dpi=300)
}
```


